External priors for the next generation of CMB experiments.

### 2016-03-12 Notes/CMBS4_priors_paper.tex

We compute the power spectra $C_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
the derivatives in \refeq{Fij_def2} using a 5 point finite difference formula:
this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives. As a consequence the differences of power spectra corresponding to different values of the parameters are big enough to ensure numerical accuracy. We also test the robustness of this calculation by varying the derivative steps in the range $2-7\%$ . The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties. 
Our philosophy is to consider as few extensions as possible. Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models. As such, the key question is: What information is needed to reliably conclude that an additional parameter is required? In our baseline model we decide to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist. So our fiducial cosmology is flat $\nu \Lambda$CDM, with assumed 
parameters from Table 2 of the \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc, supplemented by an choice of $\sum m_\nu$ $\simeq$ 85\ meV to be consistent with \cite{wu:2014}.  
We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters, in each case keeping the other parameter fixed.
The fiducial values of these are $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1$. In this same spirit, we keep the curvature fixed to zero, as the level predicted in the vast majority of inflationary models is of order $10^{-5}$, much too small to have an impact on the parameters we consider. Similarly, when we vary $N_{\rm eff}$ we adjust the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends also on the baryon density.

@@

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
Indeed, below the free-streaming scale, the matter power spectrum is suppressed in the presence of
massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
This effect is probed by CMB lensing and, since the active neutrino number densities are known in the Standard Model, it can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$. This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).

@@

\reffig{prior_w-h} shows that without any priors, CMB experiments (+BAO15) will do better than current constraints, which hover around 10\%. However, an external an external prior on $H_{0}$ two times more accurate than CMB S4 (i.e. $1\%$) can improve the CMB constraint on $w$ by a factor of two. 
This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface. However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the Universe flat). Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.

@@

Toward that goal, we investigated here how these constraints depend on external priors:
\begin{description}
\item[$\sum m_\nu$] To improve the CMB constraint on the sum of neutrino masses by a factor of two, low-redshift measurements will be needed. Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub-percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
Furthermore, if the final design of CMB S4 does not include measurements on large scales ($\ell<50$), external constraints from 21 cm experiments or all-sky CMB experiments on the optical depth $\tau$ will become very important.% powerful.
\item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$. These are most likely to come from surveys that measure a large fraction of the $z>1$ linear Fourier modes.
\item[$w$] In this case the improvement will mainly come from external priors on $H_{0}$. For example a $1\%$ prior from an external experiment will reduce the error on $w$ by a factor of two.\end{description}

>>>>

We compute the power spectra $C_{\ell}$ in \refeq{Fij_def2} using CAMB \cite{Howlett:2012mh,Lewis:1999bs} and
the derivatives in \refeq{Fij_def2} using a 5 point finite difference formula:
this high order approach allows us to use larger step-sizes around the fiducial parameters to compute derivatives. As a consequence the differences of power spectra corresponding to different values of the parameters are big enough to ensure numerical accuracy. We also test the robustness of this calculation by varying the derivative steps in the range $2-7\%$ . The constraints change by at most $10\%$ that should then be considered as a conservative estimate of numerical uncertainties. 
Our philosophy is to consider as few extensions as possible. Given the current success of the standard models of particle physics and cosmology, one of the primary goals of CMB-S4 will be to find cracks in these models. As such, the key question is: What information is needed to reliably conclude that an additional parameter is required? In our baseline model we decide to include massive neutrinos, which (i) are a small extension to the Standard Model and (ii) are known to exist. So our fiducial cosmology is flat $\nu \Lambda$CDM, with assumed 
parameters from Table 2 of the \textit{Planck} best fit \cite{planck-collaboration:2014g}, i.e. $\Omega_c h^2 = 0.12029$, $\Omega_b h^2 = 0.022068$, $A_s = 2.215\times10^{-9}$ at $k_0 = 0.05\ {\rm Mpc}^{-1}$, $n_s = 0.9624$, $\tau = 0.0925$, $H_0 = 67.11$ km/s/Mpc, supplemented by an choice of $\sum m_\nu$ $\simeq$ 85\ meV to be consistent with \cite{wu:2014}.  
We then extend the parameter space by introducing $N_{\rm eff}$ and $w$ as free parameters, in each case keeping the other parameter fixed.
The fiducial values of these are $N_{\rm eff}=3.046$ and a cosmological constant equation of state, $w=-1$. In this same spirit, we keep the curvature fixed to zero, as the level predicted in the vast majority of inflationary models is of order $10^{-5}$, much too small to have an impact on the parameters we consider. When we vary $N_{\rm eff}$ we adjust the helium abundance to its canonical value from Big Bang Nucleosynthesis, which depends also on the baryon density.

@@

Small scale structure formation is suppressed if fast-moving neutrinos comprise a significant part of the matter budget. 
Indeed, below the free-streaming scale, the matter power spectrum is suppressed in the presence of
massive neutrinos by a factor $\Delta P/P\simeq -8f_{\nu}$ , where $f_{\nu} = \Omega_{\nu} / \Omega_{m}$ is the contribution of neutrinos to the total matter density.
This effect is probed by CMB lensing and, considering the robust prediction from particle physics Standard Model for the number density of active neutrinos, it can be directly transformed into a constraint on the sum of the neutrino masses $\sum m_\nu$. This is particularly exciting because we now know that neutrinos are massive, with a lower limit $\sum m_\nu>50$ meV that emerges from oscillation experiments (see \cite{bellini:2013} and references therein).

@@

\reffig{prior_w-h} shows that without any priors, CMB experiments (+BAO15) will do better than current constraints, which hover around 10\%. However, an external prior on $H_{0}$ two times more accurate than CMB S4 (i.e. $1\%$) can improve the CMB constraint on $w$ by a factor of two. 
This is primarily due to the fact that CMB constrains $w$ through a precise measurement of the comoving distance to the last scattering surface. However that can be kept fixed while varying $w$ by accordingly changing the value of $H_{0}$ (and $\Omega_{\Lambda}$ to keep the Universe flat). Indeed in a two dimensional ($H_{0},w$) space the CMB constraint approximately lies on the region of constant large scattering surface distance as shown in \reffig{ellipse_w_H0}.

@@

Toward that goal, we investigated here how these constraints depend on external priors:
\begin{description}
\item[$\sum m_\nu$] To improve the CMB constraint on the sum of neutrino masses by a factor of two, low-redshift measurements will be needed. Indeed an external prior on $\Omega_{c}h^{2}$ and $H_{0}$ at the level attained by CMB-S4 (sub-percent) can reduce the 1-$\sigma$ uncertainty by almost a factor of two.
Furthermore, if the final design of CMB S4 does not include measurements on large scales ($\ell<50$), external constraints on the optical depth $\tau$, from 21 cm experiments or CMB experiments measuring the large
scales, will become very important.
\item[$N_{\rm eff}$] The constraint on the relativistic degrees of freedom will benefit from external priors on parameters that affect the shape of the damping tail, such as the spectral index $n_s$. These are most likely to come from surveys that measure a large fraction of the $z>1$ linear Fourier modes.
\item[$w$] In this case the improvement will mainly come from external priors on $H_{0}$. For example a $1\%$ prior from an external experiment will reduce the error on $w$ by a factor of two.\end{description}
